---
title: "Data Wrangling - Missing Values"
author: "Tom Thorpe"
date: "March 27, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Springboard Data Science Foundation Class
Exercise 01 - Data Wrangling - Refine Data

## Objective
Practice using *tidy* and *dplyr* packages to clean up data in a practice dataset.

## Exercise Results
Load the tidy and dplyr libraries for exercise.
```{r message=FALSE}
#library(devtools)
library(tidyr)
library(dplyr)
# install.packages("gdata") needed for read.xls function
# library(gdata) - instructions were to convert to csv before loading, so not needed.
```

### Identify input and output files
The practice data set was downloaded into a file called "titanic3.xls", resaved as
"titanic_original.csv" and
the cleaned data will be stored in a file named, "titanic_clean.csv".
```{r}
infile = "C:/Users/Tom/git/datasciencefoundation/DataWrangleExer02/titanic_original.csv"
outfile = "C:/Users/Tom/git/datasciencefoundation/DataWrangleExer02/titanic_clean.csv"  
```

### Read CSV file into a local dataframe
```{r}
# titanic <- read.xls(infile) %>% tbl_df # instructions were to convert to csv before loading
#titanic <- read.csv(file=infile, header=TRUE, sep=",") %>% tbl_df()
titanic1 <- read.csv(file=infile, header=TRUE, sep=",") 
titanic <- tbl_df(titanic1)
```

Lets see what the data looks like:
```{r}
str(titanic)
titanic
```
I noticed that there are three additional variables in the data set than in the data description
pointed to by the exercise URL, 
*boat*, *body* and *home.dest*. I found the full description of the titanic3.xls data set here
<http://biostat.mc.vanderbilt.edu/wiki/pub/Main/DataSets/titanic3info.txt>. 
The *boat* is the lifeboat, *body* is the body identification number, and *home.dest* is the home
destination of the passenger.

### Default Missing *embarked* data
First fill in any missing data in the *embarked* column with 'S' to represent Southampton.
First, Check to see how many are blank.
```{r}
table(titanic$embarked)
```
There are two empty *embarked* values. Make the updates.
```{r}
titanic$embarked[titanic$embarked==""|titanic$embarked==" "] <- "S"
#titanic$embarked[grepl("^ |! {0}",titanic$embarked)] <- "S" # couldn't figure out a regular expression
```
Check the results of the change.
```{r}
table(titanic$embarked)
```

### Populate missing *age* data
Populate the missing data with the mean of the other rows with age data.
Look at the age data before populating.
```{r}
table(titanic$age)
summary(titanic$age)
filter(titanic,is.na(age)) %>% count
select(titanic,name,age) %>% print(n=18)
```
Notice that the 16th entry has NA for age. Set the empty age data to the mean of the non-empty age.
```{r}
meanAge= mean(titanic$age,na.rm=TRUE)
medianAge= median(titanic$age,na.rm=TRUE)
# meanAge= mean(titanic$age[titanic$age>=0],na.rm=TRUE)
meanAge
medianAge
meanAge<-trunc(meanAge) # truncate the age since only age less than 1 have fractional values.
#titanic$age[titanic$age==NA] <- meanAge
titanic$age[is.na(titanic$age)] <- meanAge

#Check the results
select(titanic,name,age) %>% print(n=18)
filter(titanic,is.na(age)) %>% count
```
How else could the missing age data be populated? The median is an alternative measure that may be 
closer to a representative age. The median in this case is 28. The mean is 29.88 rounded to 30. I believe that since the median is less than the mean, there are more younger people than old people on the ship.
I would tend to use the median over the mean in this case. Somehow it just feels better. Maybe I will 
have a better answer after taking the statistics part of the course.

### Populate missing Lifeboat data
There is missing data for the lifeboat column.

Does it make sense to fill in the cabin numbers with a value? To me, it does not make sense to 
try to come up with an estimate of the cabin number. How could any meaningful data be created?
It makes sense to set the missing data to "NONE".

What does a missing value mean here? Missing data could mean the records for the cabin assignment
were lost or the people were assigned to a big dormitory area instead of a room. But there was 
probably more than one dormitory and the specific dormitory is unknown.
Either way, the best that can be done is to do as the assignment suggests and assign missing
data the value of "NONE".

Update the missing data in the lifeboat data with "NONE". I tried using these commands

  **titanic\$boat[titanic\$boat==""] <- as.factor("NONE")**
  
  **titanic\$boat[titanic\$boat==""] <- "NONE"**

But both resulted in the following warning:

   Warning in `[<-.factor`(`*tmp*`, titanic$boat == "", value =
   structure(c(13L, : invalid factor level, NA generated
   
So I did some research and converted the *boat* column from a factor to character data. But I don't understand why the column was a factor and not character data to begin with. I need to research this.  
```{r}
filter(titanic, boat =="") %>% count()
titanic$boat <- as.character.factor (titanic$boat)
titanic$boat[titanic$boat==""] <- "NONE"
#Check the results
select(titanic,name,boat) %>% print(n=15)
filter(titanic, boat =="") %>% count()
```
### Create has_cabin_number column
Create a *has_cabin_number* column with a value of 1 if the *cabin* column is non-blank.
```{r}
titanic <- mutate(titanic,has_cabin_number=ifelse(cabin!="",1,0))
# Check results
select(titanic,name,cabin,has_cabin_number) %>% print(n=15)
glimpse(titanic)
```

### Save the cleaned data
```{r}
write.csv(titanic, file=outfile,row.names=FALSE)
```
That concludes the exercise.
